Legacy Data Confounds Modern Genomics Studies

### 2018-06-19 legacy.tex

\maketitle
\begin{abstract}
In investigating a population genetics signal, we noticed some discrepancies between the Japanese samples from the 1000 Genomes Project (1kGP) and a more recent high quality dataset.
We found that the variants causing the signal were only present in individuals with low average sequencing quality.
We identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ single nucleotide polymorphisms (SNPs) that were associated to quality in the 1kGP dataset Japanese individuals.
These variants are present in nearly all of the populations from the 1kGP. 
We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.
Some of these variants are being imputed onto genotype data and reach genome wide significance in recent publications.
\todo{update values for all populations}
\end{abstract}

@@

Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.
While trying to follow up on this observations in a larger Japanese cohort individuals, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups.
Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.

When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}, suggesting a technical artifact rather than a population structure effect.
Regressions against different quality metrics provided by the 1kGP revealed that mean quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
Individuals with low-quality data have elevated rates of the *AC${\rightarrow}$*CC signature. 
\sgcomment{supplementary figure of correlations}
Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.

To identify SNPs that are associated with low quality, we performed a genome-wide association study (GWAS)  \ref{SFS}. 
We used the mean quality of mapped bases of each individual as the association phenotype.
Thus, variants that are only found in individuals with low quality data will display significant associations to quality. 
We were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.  
This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
The variants that are associated to the quality of mapped bases have a 10 fold enrichment in *AC${\rightarrow}$*CC mutations.

The GWAS has a power limit of 5\% minor allele frequency (MAF), many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality.
The *AC${\rightarrow}$*CC mutational signal is limited to individuals with low quality sequencing data.
Therefore, the removal of individuals with average qualities below 30 successfully removes the signal.

By taking advantage of the larger, higher quality dataset, we were also able to get rid of this signal by filtering all the variants above 1\% in the 1kGP Japanese but missing from the \todo{NAG} data.
This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.

@@

	\subsection{Overlap of significant SNPs}

We applied the same GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
We found that 25 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
\todo{Supplemental Figures : all the GWAS results for each population}
Comparing the results of each independent GWAS, we were able to identify over 3625 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.

@@

\begin{center}
 \begin{tabular}{||c c c c||} 
 \hline
Author & Year & Journal & RSID \\ [0.5ex] 
 \hline\hline
Comuzzie AG &	2012	 &PLoS One &	rs9469457\\ 
 \hline
Smith JG &	2012	 & Circ Cardiovasc Genet	& rs7729539\\ 
 \hline
Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648\\ 
 \hline
Zeng Z	& 2014 &	Caries Res	& rs10495332\\ 
 \hline
Park SL & 	2015	 & PLoS One	& rs113626797\\ 
 \hline
Parmar PG &	2016	 & Circ Cardiovasc Genet	& rs12370001\\ 
 \hline
Spada J & 	2016	 & J Sleep Res &	rs12452225\\ 
 \hline
Astle WJ	& 2016 &	Cell	 & rs533229216\\ 
 \hline
 Ahola-Olli AV &	2017 &	Am J Hum Genet &	rs76842834\\ 
 \hline
Liu TC	& 2017 &	JCI Insight &	rs12481514\\ 
 \hline
  McKay JD & 2017 &	Nature Genetics & 	rs13191445\\ 
 \hline
Zhou H &	2017	 & JAMA Psychiatry &	rs11179306\\ 
 \hline
Tian C	& 2017	 &Nature Communications	 & rs67491322\\ 
 \hline
Qian DC	& 2017 &	BMC Genomics &	rs77576211\\ 
 \hline
Michailidou K	& 2017	 & Nature &	rs76092474\\ 
 \hline
& & & rs80274284, rs201130852,\\ 
Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,\\ 
& & & rs201255786, rs201761909\\ 
 \hline
 \end{tabular}
\end{center}

@@

\section{Methods}
\subsection{Metadata}
The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
Each BAS file has metadata regarding each sequencing event for each sample. 
If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample. 
The dates of the sequencing are only used to plot Figure. \todo{Average Quality of mapped bases: How was it calculated?}

@@

Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1kGP. We used the average quality of mapped bases per individual as the phenotype for the analysis. We also controlled for population structure by including the first 4 principle components of a PCA of each population using genotype data. \todo{NOT DONE YET}

@@

\bibliography{/Users/luke/Dropbox/Work/PHD/BibTeX/Legacy.bib}

>>>>

\maketitle
\begin{abstract}
In investigating an unusual population genetics signal we identified a strong and previously unreported batch effect in the 1000 Genomes Project data.
We also found that this batch effect led to spurious results in a number of recent publications. 
We identified spurious mutations by correlating mutations with data quality metrics.
We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.
Some of these variants are being imputed onto genotype data and reach genome wide significance in recent publications.
\end{abstract}

@@

Harris and Pritchard reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.
While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
This is unexpected given the similarity of the two populations, and suggests a technical artifact rather than a population structure effect. 
Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.

When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}.
Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP. 
Individuals with low-quality data showing elevated rates of the signature. \ref{PC1_Correlation}
Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality \ref{SFS}. 
Using the linear regression as implemented in PLINK, we were able to identify $625$ snips with $p < 10^{-8}$ and $1048$ snps with $ p < 10^{-6}.$ 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., the traditional genome-wide significance threshold might not be a good guide for filtering).  
 This GWAS included 104 Japanese individuals, \todo{000} of which were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
The variants that are associated to the quality of mapped bases have a 10 fold enrichment in *AC${\rightarrow}$*CC mutations.

The signal is enriched in GWAS snps, but residual signal remains because our GWAS has a power limit of 5\% minor allele frequency (MAF) resulting in many rare alleles found in individuals with low-quality data remaining unidentifiable using association techniques.
For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality.
The *AC${\rightarrow}$*CC mutational signal is limited to individuals with low quality sequencing data.
Therefore, the removal of individuals with average qualities below 30 successfully removes the signal.
While the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.

@@

	\subsection{Overlap of significant SNPs}
We applied the same GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
We found that 25 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
\todo{Supplemental Figures : all the GWAS results for each population}
Comparing the results of each independent GWAS, we were able to identify over 3625 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.

@@

\begin{center}
 \begin{tabular}{||c c c c c||} 
 \hline
Author & Year & Journal & RSID & PMID \\ [0.5ex] 
 \hline\hline
Comuzzie AG &	2012	 &PLoS One &	rs9469457&23251661\\ 
 \hline
Smith JG &	2012	 & Circ Cardiovasc Genet	& rs7729539&23166209\\ 
 \hline
Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\ 
 \hline
Zeng Z	& 2014 &	Caries Res	& rs10495332&24556642\\ 
 \hline
Park SL & 	2015	 & PLoS One	& rs113626797&26053186\\ 
 \hline
Parmar PG &	2016	 & Circ Cardiovasc Genet	& rs12370001&26969751\\ 
 \hline
Spada J & 	2016	 & J Sleep Res &	rs12452225&27126917\\ 
 \hline
Astle WJ	& 2016 &	Cell	 & rs533229216&27863252\\ 
 \hline
 Ahola-Olli AV &	2017 &	Am J Hum Genet &	rs76842834&27989323\\ 
 \hline
Liu TC	& 2017 &	JCI Insight &	rs12481514&28352666\\ 
 \hline
  McKay JD & 2017 &	Nature Genetics & 	rs13191445&28604730\\ 
 \hline
Zhou H &	2017	 & JAMA Psychiatry &	rs11179306&	29071344\\ 
 \hline
Tian C	& 2017	 &Nature Communications	 & rs67491322&28928442\\
 \hline
Qian DC	& 2017 &	BMC Genomics &	rs77576211&28927378\\ 
 \hline
Michailidou K	& 2017	 & Nature &	rs76092474&29059683\\ 
 \hline
& & & rs80274284, rs201130852,&\\
Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
& & & rs201255786, rs201761909&\\
 \hline
 \end{tabular}
\end{center}

@@

\section{Methods}
\subsection{Metadata}
The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
Each BAS file has metadata regarding each sequencing event for each sample. 
If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample. 
The dates of the sequencing are only used to plot Figure.

@@

Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1kGP. We used the average quality of mapped bases per individual as the phenotype for the analysis. We controlled for population structure by including the first 4 principle components of a PCA of each population with available genotype data, however, considering that population structure did not affect the number of variants associated to quality, and that not all populations had genotype data readily available, we did not include these in the final analysis.

@@

\subsection{Supplementary Figures}
\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{PC1_Correlation.jpg}
\caption{Regressions against mean quality per mapped base pair per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.  }
 \label{PC1_Correlation}
\end{figure}

\bibliography{Legacy.bib}
